---
title: "int64: 64-bit Integer Vectors in R"
author: "Romain Francois"
date: '`r Sys.Date()`'
output:
  html_document:
  fig_caption: yes
  theme: cerulean
  toc: yes
  toc_depth: 2
vignette: >
  %\VignetteIndexEntry{64-bit Integer Vectors in R}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r echo=FALSE}
knitr::opts_chunk$set(
    message = FALSE,
    warning = FALSE,
    error = FALSE,
    tidy = FALSE,
    cache = FALSE
)
prettyVersion <- packageDescription("int64")$Version
prettyDate <- format(Sys.Date(), "%B %e, %Y")
require( int64 )
```

## 1. Introduction

The `int64` package adds 64 bit integer vectors to `R`. 
The package provides the `int64` and `uint64` classes for
signed and unsigned integer vectors. This project has been 
sponsored by the Google Open Source Programs Office. 

## 2. Background

Integers in `R` are represented internally as 32 bit `int`. 
Aplications now require larger ranges of values to represent large quantities. 
This package exposes C++ types `int64_t` and `uint64_t`
to `R` for this purpose. The table below shows the limits of these
types. 

```{r, echo = FALSE}
d <- data.frame(a = c('`int`', '`int64_t`', '`uint64_t`'))
d$b <- c('`integer`', '`int64`', '`uint64`')
d$c <- c(numeric_limits("integer")[1L], as.character(numeric_limits("int64")[1L]), as.character(numeric_limits("uint64")[1L]))
d$d <- c(numeric_limits("integer")[2L], as.character(numeric_limits("int64")[2L]), as.character(numeric_limits("uint64")[2L]))
names(d) <- c("C++ type", "R type", "min", "max")
knitr::kable(d, caption = "Numeric limits of integer types")
```

## 3. Usage

This section shows a few examples on how to use the package. 

```{r}
# create a new int64 vector
x <- int64( 4 )
```

```{r}
# set a subset of values
x[1:2] <- 1:2 # via integers
x[3:4] <- c("123456789123456", "-9876543219876") # ... or characters
x
```

```{r}
# convert integer or character vectors into int64 vectors
x <- as.int64( 1:6 )
x
y <- as.int64( c("-1234", "1234" ) )
y
```

```{r}
# create a data frame with a column of int64
df <- data.frame( a = 1:4 )
df$y <- as.int64( 1:4 )
df
```

## 4. The `int64` and `uint64` classes

### 4.1. Class representation

Both `int64` and `uint64` are represented as lists of pairs of
integers. 

```{r}
str( as.int64( 1:2 ) )
```

Each `int64` or `uint64` number is represented as a couple of 32 bit integers. 
Internally, the C++ code goes back and forth between the native representation
of these numbers as C++ data types (`int64_t` and `uint64_t`)
and their representation as couples of 32 bit integers by splitting the 
64 bits. 

For example, the `int64_t` value (-123) is represented in memory as: 

```{r}
binary( as.int64( "-123" ))@data
```

These 64 bits are split into the two following chunks: 

```{r}
first.int <- as.int64( "-123" )[[1L]][1L]
first <- binary( first.int)@data
second.int <- as.int64( "-123" )[[1L]][2L]
second <- binary( second.int )@data
```

```{r, echo = FALSE}
d <- data.frame(a = first, b = second)
names(d) <- c("First chunk", "Second chunk")
knitr::kable(d)
```

The R representation of -123 is therefore composed by the two integers
whose binary representation is above, i.e (\Sexpr{first.int},\Sexpr{second.int}).
    This representation has been chosen against other alternatives to 
allow these key requirements: 

- Data must be serializable
- `int64` and `uint64` vectors have to be usable of  columns of data frames. 
- The `int64` and `uint64` types must supposrt missing values (`NA`)
\end{itemize}
            
### 4.2. Creating new vectors

The functions `int64` and `uint64` can be used to create
new vectors of signed or usigned 64 bit integers of the given length. These
functions are similar to the usual `R` functions `numeric`, 
`integer`, etc ...

```{r}
int64(3)
uint64(10)
```

### 4.3. Converting integer or character vectors

The functions `as.int64` and `as.uint64` can be used
to convert `integer` or `character` vectors into signed or 
unsigned 64 bit integers. 

```{r}
as.int64( 1:4 )
as.uint64( c("123456789", "987654321987654321" ) )
```
   
Internally `integer` vectors are converted using a reguar cast, and
`character` vectors are converted using the `C` function
`atol`. 

### 4.4. Subsetting

Extracting or setting subsets from a `int64` or `uint64`
vector is similar to other vector classes in R. 

```{r}
x <- as.int64( 1:4 )
x[1:2]
x[3:4] <- 5:6
x
```

### 4.5. Arithmetic operations

The `Arith` group generic is implemented for classes `int64`
and `uint64`. 

```{r}
x <- as.int64( 1:4 )
x + 1L
x - 1:2
x * x
x / 2L
x %% 2L
x %/% 2L
```


### 4.6. Logical operations

The `Compare` group generic is implemented
for classes `int64` and `uint64`. 

```{r}
x <- as.int64( 1:5 )
x < 3L
x > 6L - x
x != 3L
x == 4L
x <= 3L
x >= 5L
```

### 4.7. Summary operations

The `Summary` group generic is implemented
for classes `int64` and `uint64`. 

```{r}
x <- as.int64( 1:5 )
min( x )
max( x )
range( x )
prod( x )
sum( x )
any( x )
all( x )
```
         

### 4.8. Binary representation

The `binary` generic function shows the bit representation
of `numeric`, `integer`, `int64` and `uint64`. 
                             
```{r}
binary( 1:4 ) # integer
binary( c(1.2, 1.3) ) # numeric
binary( as.int64( 1:4 ) ) # signed 64 bit integer (int64)
binary( as.uint64( 1:4 ) ) # unsigned 64 bit integer (uint64)
```

## 5. Numeric limits and missing values

The `numeric_limits` function gives the limits for 
types `integer`, `int64`, `uint64`. 

```{r}
numeric_limits( "integer" )
numeric_limits( "int64" )
numeric_limits( "uint64" )
```

`int64` and `uint64` classes support missing values using the same mechanism as
R uses for integer vectors. 

For signed 64 bit integer vectors (`int64`), `NA` is represented by the 
value $-2^{63}$, hence the range of acceptable values is 

$$[-2^{63}+1,2^{63}-1]$$ 

For unsigned 64-bit integer vectors (`uint64`), `NA` is represented by the
value $2^{64}-1$, hence the range of acceptable values is

$$[0,2^{64}-2]$$

### 6. Reading 64 bit integers from files

The `int64` implements the necessary methods so that `read.csv`
can read signed and unsigned 64 bit integers from files. 

```{r}
tf <- tempfile()
df <- data.frame( x = 1:10, y = 1:10, z = 1:10 )
write.table( df, tf, sep = ",", row.names = FALSE )
df <- read.csv( tf, colClasses = c("integer", "int64", "uint64" ) )
df
sapply( df, class )
```

